---
title: "Code supporting the examples in the paper _Communicating results in trials with multiple hypotheses or adaptive design features_"
author: "Marcel Wolbers, on behalf of the author team"
date: 'Last modification: `r Sys.setlocale("LC_TIME", "C"); format(Sys.Date(), "%d %b %Y")`'
format:
  html:
    toc: true
    toc-depth: 2
    toc-title: Contents
    toc-location: left
    number-sections: true
    number-depth: 2
bibliography: stat.bib
---

# Background

Code for the paper _Communicating results in trials with multiple hypotheses or adaptive design features_, Asikanius et al. [-@asikanius_26], written by an author team of statisticians from regulators, academia, and industry.

# Purpose of this document

This R Quarto file provides easy accessible code to compute all the quantities in the paper. The github repository where this document is hosted is available [here](https://github.com/oncoestimand/complex_designs_communication).

# Setup {.tabset .tabset-fade .tabset-pills}

```{r setup, message= FALSE, warning = FALSE}
# packages
library(MASS)
library(gMCPLite) # graphical multiple testing procedures 
library(rpact)    # group-sequential and adaptive trials

# Source functions for calculating bias-adjusted estimators in GS trials via simulation
source("biasCorrectGroupSeqSurvival.r")
```

# Example 1: A randomized trial in early Alzheimer’s disease with hierarchical testing {-}

In this section we reproduce numbers in Table 1 of the paper.

```{r ex1_unadj}
# ----------------------------------------
# Unadjusted trial results 
# ----------------------------------------

alpha <- 0.05
za <- qnorm(1 - alpha / 2)
endpoints <- c("CDR-SB", "ADAS-Cog13", "FAQ")
estimate <- c(-0.4, -1.3, -1.1)
se <- c(0.6, 3, 1.6) / (2 * za)

# one-sided p-value
p_values_less <- pnorm(estimate / se)
p_values_greater <- 1 - p_values_less

# per comparison two-sided CI and p-values 
data.frame(
  endpoints,
  estimate, 
  ci_lower_unadj = estimate - za * se, 
  ci_upper_unadj = estimate + za * se, 
  p_value_2sided_unadj  = 2 * pmin(p_values_greater, p_values_less),
  p_value_2sided_unadj_r  = round(2 * pmin(p_values_greater, p_values_less), 2)
  )
```

```{r ex1_adj}
# ----------------------------------------
# Adjusted trial results 
# ----------------------------------------

# Define transition matrix and weights for hierarchical testing 
# (as a graphical multiple comparison procedure)
transitions <- rbind(
  c(0, 1, 0),
  c(0, 0, 1),
  c(0, 0, 0)
)

weights <- c(1, 0, 0)
hierarchical_graph <- matrix2graph(transitions, weights)

# ----------------------------------------
# plot graph (one-sided at level 0.025)
# ----------------------------------------
hGraph(
  nHypotheses = 3,
  nameHypotheses = endpoints,
  alphaHypotheses = c(alpha / 2, 0, 0),
  m = transitions
)

# ----------------------------------------
# adjusted p-values 
# ----------------------------------------

# direction "greater"
result_greater <- gMCP(
  graph = hierarchical_graph, 
  pvalues = p_values_greater, 
  alpha = alpha / 2
)

# one-sided 
result_greater@adjPValues 

# direction "less"
result_less <- gMCP(
  graph = hierarchical_graph, 
  pvalues = p_values_less, 
  alpha = alpha / 2
)

# one-sided 
result_less@adjPValues 

# ----------------------------------------
# simultaneous confidence intervals 
# ----------------------------------------

# One-sided 97.5% CI, direction "greater"
ci_results_greater <- simConfint(
  object = hierarchical_graph, 
  pvalues = p_values_greater, 
  confint = "normal",       
  estimates = estimate, 
  alpha = alpha / 2,
  alternative = "greater" 
)

ci_results_greater

# One-sided 97.5% CI, direction "less"
ci_results_less <- simConfint(
  object = hierarchical_graph, 
  pvalues = p_values_less, 
  confint = "normal",       
  estimates = estimate, 
  alpha = alpha / 2,
  alternative = "less" 
)

ci_results_less

# Unadjusted two-sided CI (intersection of the two one-sided CI) 
# and two-sided p-value
data.frame(
  endpoints, 
  estimate = estimate,
  ci_lower_adj = ci_results_greater[, "lower bound"], 
  ci_upper_adj = ci_results_less[, "upper bound"], 
  p_value_2sided_adj  = 2 * pmin(result_greater@adjPValues, result_less@adjPValues),
  p_value_2sided_adj_r  = round(2 * pmin(result_greater@adjPValues, result_less@adjPValues), 2)
)
```

# Example 2: A group-sequential trial with hierarchical testing in oncology {-}

In this section we reproduce numbers in Table 2 of the paper.

## Results for progression-free survival (PFS) {-}

### Trial design 

```{r ex2_pfs_design}
# ----------------------------------------
# primary endpoint: PFS
# ----------------------------------------

# 230 events total -> 90% power for HR=0.65
# IA after 152 events (66%); 
# observed at IA:
#   - log(HR) = log(0.61)
#   - SE = sqrt(4 / 152)

alpha <- 0.05
za <- qnorm(1 - alpha / 2)
beta <- 0.1
d_int <- 152
d_fin <- 230
n <- 500
hr <- 0.65

# Group-sequential design
infoRates1 <- c(d_int / d_fin, 1)

d1 <- getDesignGroupSequential(
  informationRates = infoRates1,typeOfDesign = "asOF",
  sided = 2, alpha = alpha, beta = beta)
summary(d1)

s1 <- getPowerSurvival(
  d1, maxNumberOfSubjects = n, maxNumberOfEvents = d_fin, 
  median1 = 24, hazardRatio = hr)
summary(s1)
```

### Naive inference {-}

```{r ex2_pfs_naive}
# ----------------------------------------
# naive inference
# ----------------------------------------

# observed at interim
log_est1 <- log(0.61)
se1 <- sqrt(4 / 152)

log_est1_lower <- log_est1 - za * se1
log_est1_upper <- log_est1 + za * se1

# estimate and CI
round(exp(c(log_est1, log_est1_lower, log_est1_upper)) , 2)

# 2-sided p-value
round(2 * pnorm(log_est1 / se1) , 3)
```

### Inference based on stagewise ordering {-}

```{r ex2_pfs_stagewise}
# ----------------------------------------
# Median unbiased estimator, CI, p-value
# all based on stage-wise ordering
# repeated CI and p-value with rpact
# ----------------------------------------
#
# !! For rpact v4.30, the "Final p-value" (i.e. p-value based on stage-wise ordering) 
#    is reported  as 0.9988 in the output below but it should rather be 
#    1-0.9988 = 0.0012 instead. 
#    [See https://github.com/rpact-com/rpact/issues/114 which will hopefully be 
#    fixed soon]
r1 <- getDataSet(
  overallEvents = c(d_int),
  overallLogRanks = c(log_est1 / se1),
  overallAllocationRatio = 1
)

adj_r1 <- getAnalysisResults(
  design = d1, dataInput = r1
)

summary(adj_r1)
```

### Unconditional (global) bias adjusted estimator via simulation  {-}

```{r ex2_global}
boundaryEffectScaleLower <- s1$criticalValuesEffectScaleLower[,1]
boundaryEffectScaleUpper <- s1$criticalValuesEffectScaleUpper[,1]

nsim <- 1E6
seed <- 12

# estimator
uncond_est <- biasCorrectGroupSeqSurvival(
  hazardRatioMLE = exp(log_est1), 
  stopstage = 1, 
  type = "global",
  n = nsim, 
  informationRates = infoRates1, 
  maxNumberOfEvents = d_fin, 
  allocationRatioPlanned = 1,
  boundaryEffectScaleLower = boundaryEffectScaleLower,
  boundaryEffectScaleUpper = boundaryEffectScaleUpper,
  seed = seed)

# crude lower CI by simply adjusting the naive lower CI
# (as proposed by Pinheiro & deMets)
uncond_low <- biasCorrectGroupSeqSurvival(
  hazardRatioMLE = exp(log_est1_lower), 
  stopstage = 1, 
  type = "global",
  n = nsim, 
  informationRates = infoRates1, 
  maxNumberOfEvents = d_fin, 
  allocationRatioPlanned = 1,
  boundaryEffectScaleLower = boundaryEffectScaleLower,
  boundaryEffectScaleUpper = boundaryEffectScaleUpper,
  seed = seed)

# crude upper CI by simply adjusting the naive upper CI
# (as proposed by Pinheiro & deMets)
uncond_up <- biasCorrectGroupSeqSurvival(
  hazardRatioMLE = exp(log_est1_upper), 
  stopstage = 1, 
  type = "global",
  n = nsim, 
  informationRates = infoRates1, 
  maxNumberOfEvents = d_fin, 
  allocationRatioPlanned = 1,
  boundaryEffectScaleLower = boundaryEffectScaleLower,
  boundaryEffectScaleUpper = boundaryEffectScaleUpper,
  seed = seed)

round(c(uncond_est, uncond_low, uncond_up), 2)
```

### Conditional bias adjusted estimator via simulation {-}

```{r ex2_conditional}
# estimator
cond_est <- biasCorrectGroupSeqSurvival(
  hazardRatioMLE = exp(log_est1), 
  stopstage = 1, 
  type = "stagewise",
  n = nsim, 
  informationRates = infoRates1, 
  maxNumberOfEvents = d_fin, 
  allocationRatioPlanned = 1,
  boundaryEffectScaleLower = boundaryEffectScaleLower,
  boundaryEffectScaleUpper = boundaryEffectScaleUpper,
  seed = seed)

# crude lower CI by simply adjusting the naive lower CI 
cond_low <- biasCorrectGroupSeqSurvival(
  hazardRatioMLE = exp(log_est1_lower), 
  stopstage = 1, 
  type = "stagewise",
  n = nsim, 
  informationRates = infoRates1, 
  maxNumberOfEvents = d_fin, 
  allocationRatioPlanned = 1,
  boundaryEffectScaleLower = boundaryEffectScaleLower,
  boundaryEffectScaleUpper = boundaryEffectScaleUpper,
  seed = seed)

# crude upper CI by simply adjusting the naive upper CI 
cond_up <- biasCorrectGroupSeqSurvival(
  hazardRatioMLE = exp(log_est1_upper), 
  stopstage = 1, 
  type = "stagewise",
  n = nsim, 
  informationRates = infoRates1, 
  maxNumberOfEvents = d_fin, 
  allocationRatioPlanned = 1,
  boundaryEffectScaleLower = boundaryEffectScaleLower,
  boundaryEffectScaleUpper = boundaryEffectScaleUpper,
  seed = seed)

round(c(cond_est, cond_low, cond_up), 2)
```

## Results for overall survival (OS) {-}

```{r ex2_OS}
# ----------------------------------------
# secondary endpoint: OS
# ----------------------------------------
# 
# 233 events -> 80% power for HR = 0.69
# IA at 50% and 75% IF
# Results for IA after 117 events (50%):
#   log(HR) = log(0.67)
#   SE = sqrt(4 / 117)

# ----------------------------------------
# trial design
# ----------------------------------------

alpha <- 0.05
za <- qnorm(1 - alpha / 2)
beta <- 0.2
hr_os <- 0.69
infoRates2 <- c(0.5, 0.75, 1)

# number of events
d2 <- getDesignGroupSequential(informationRates = infoRates2,
                                      typeOfDesign = "asOF", sided = 2, 
                                      alpha = alpha)

ss_os <- getSampleSizeSurvival(design = d2, median2 = 100, 
                               hazardRatio = hr_os,
                               maxNumberOfSubjects = 2000)

# number of events
d_os <- ceiling(as.vector(ss_os$eventsPerStage))
d_os
d_int_os <- d_os[1]

# ----------------------------------------
# naive inference
# ----------------------------------------
log_est2 <- log(0.67)
se2 <- sqrt(4 / d_int_os)

# estimate and CI
round(exp(log_est2 + c(0, -1, 1) * za * se2) , 2)

# 2-sided p-value
round(2 * pnorm(log_est2/se2) , 2)

# ----------------------------------------
# Repeated CI and p-value
# ----------------------------------------
r2 <- getDataSet(
  overallEvents = c(d_int_os),
  overallLogRanks = c(log_est2 / se2),
  overallAllocationRatio = 1
)

adj_r2 <- getAnalysisResults(
  design = d2,
  dataInput = r2
)

summary(adj_r2)
```

# Example 3: A three-arm trial of two active doses versus placebo in adults with obesity {-}

In this section we reproduce numbers in Table 3 of the paper.

## Unadjusted trial results

```{r ex3_unadj}
# ----------------------------------------
# Unadjusted trial results 
# ----------------------------------------

alpha <- 0.05
za <- qnorm(1 - alpha / 2)

endpoints <- c("WL26_D1", "WL26_D2", "WL52_D1", "WL52_D2")
estimate <- c(-10, -5.1, -15, -15)
se <- c(10, 10, 10, 10) / (2 * za)

# one-sided p-value
p_values_less <- pnorm(estimate / se)
p_values_greater <- 1 - p_values_less

# per comparison two-sided CI and p-values 
data.frame(
  endpoints,
  estimate, 
  ci_lower_unadj = estimate - za * se, 
  ci_upper_unadj = estimate + za * se, 
  p_value_2sided_adj = 2 * pmin(p_values_greater, p_values_less),
  p_value_2sided_adj_r = format.pval(2 * pmin(p_values_greater, p_values_less), 2)
  )
```

## Results including data of Week 26 & 52

```{r ex3_adj}
# ----------------------------------------
# Adjusted trial results 
# including week 26 & 52 data
# ----------------------------------------

# transition matrix and weights of the graphical multiple comparisons procedure 
transitions <- rbind(
  c(0, 0.5, 0.5, 0),
  c(0.5, 0, 0, 0.5),
  c(0, 1, 0, 0),
  c(1, 0, 0, 0)
)

weights <- c(0.5, 0.5, 0, 0)
three_arm_graph <- matrix2graph(transitions, weights)

# plot graph (one-sided at overall level 0.025)
alpha <- 0.025
hGraph(
  nHypotheses = 4,
  nameHypotheses = endpoints,
  alphaHypotheses = c(alpha / 2, alpha / 2, 0, 0),
  m = transitions
)

# ----------------------------------------
# adjusted p-values 
# ----------------------------------------

# direction "greater"
result_greater <- gMCP(
  graph = three_arm_graph, 
  pvalues = p_values_greater, 
  alpha = alpha
)

# direction "less"
result_less <- gMCP(
  graph = three_arm_graph, 
  pvalues = p_values_less, 
  alpha = alpha
)

# ----------------------------------------
# simultaneous confidence intervals 
# ----------------------------------------

# One-sided 97.5% CI
# direction "greater"
ci_results_greater <- simConfint(
  object = three_arm_graph, 
  pvalues = p_values_greater, 
  confint = "normal",       
  estimates = estimate, 
  alpha = alpha,
  alternative = "greater" 
)

# One-sided 97.5% CI
# direction "less"
ci_results_less <- simConfint(
  object = three_arm_graph, 
  pvalues = p_values_less, 
  confint = "normal",       
  estimates = estimate, 
  alpha = alpha,
  alternative = "less" 
)

# ----------------------------------------
# unadjusted two-sided CI (intersection of the two one-sided CI) 
# and p-values, all two-sided
# ----------------------------------------

data.frame(
  endpoints, 
  estimate = estimate,
  ci_lower_adj = round(ci_results_greater[, "lower bound"], 1), 
  ci_upper_adj = round(ci_results_less[, "upper bound"], 1), 
  p_value_2sided_adj = 2 * pmin(result_greater@adjPValues, result_less@adjPValues),
  p_value_2sided_adj_r = format.pval(2 * pmin(result_greater@adjPValues, result_less@adjPValues), 2)
)
```
## Trial results adjusting for doses, only Week 26

```{r ex3_adj_w26}
# ----------------------------------------
# Adjusted trial results (including only week 26}
# ----------------------------------------

endpoints <- c("WL26_D1", "WL26_D2")
estimate <- c(-10, -5.1)
se <- c(10, 10) / (2 * za)

p_values_less <- pnorm(estimate / se)
p_values_greater <- 1 - p_values_less

# transition matrix and weights of the graphical multiple comparisons procedure 
transitions <- rbind(
  c(0, 0.5),
  c(0.5, 0)
)

weights <- c(0.5, 0.5)
three_arm_graph_w26 <- matrix2graph(transitions, weights)

# plot graph (one-sided at overall level 0.025)
alpha <- 0.025
hGraph(
  nHypotheses = 4,
  nameHypotheses = endpoints,
  alphaHypotheses = c(alpha / 2, alpha / 2),
  m = transitions
)

# ----------------------------------------
# adjusted p-values 
# ----------------------------------------

# direction "greater"
result_greater <- gMCP(
  graph = three_arm_graph_w26, 
  pvalues = p_values_greater, 
  alpha = alpha
)

# direction "less"
result_less <- gMCP(
  graph = three_arm_graph_w26, 
  pvalues = p_values_less, 
  alpha = alpha
)

# ----------------------------------------
# simultaneous confidence intervals 
# ----------------------------------------

# One-sided 97.5% CI, direction "greater"
ci_results_greater <- simConfint(
  object = three_arm_graph_w26, 
  pvalues = p_values_greater, 
  confint = "normal",       
  estimates = estimate, 
  alpha = alpha,
  alternative = "greater" 
)

# One-sided 97.5% CI, direction "less"
ci_results_less <- simConfint(
  object = three_arm_graph_w26, 
  pvalues = p_values_less, 
  confint = "normal",       
  estimates = estimate, 
  alpha = alpha,
  alternative = "less" 
)

# ----------------------------------------
# Unadjusted CI (intersection of the two one-sided CI) and 
# p-value, all two-sided
# ----------------------------------------

data.frame(
  endpoints, 
  estimate = estimate,
  ci_lower_adj = ci_results_greater[, "lower bound"], 
  ci_upper_adj = ci_results_less[, "upper bound"], 
  p_value_2sided_adj  = 2 * pmin(result_greater@adjPValues,result_less@adjPValues),
  p_value_2sided_adj_r  = format.pval(2 * pmin(result_greater@adjPValues,result_less@adjPValues), 2)
)
```

# Appendix: R functions used to calculate bias-adjusted estimators in Example 2 {-}

```{r, file = "biasCorrectGroupSeqSurvival.r", eval = FALSE, echo = TRUE}
```

